Proof-Reconstruction in Propositional Logic

### 2017-10-20 paper/sections/abstract.tex
[ paper ] restructuring the paper :S.

\begin{abstract}
We describe a syntactical proof-reconstruction
approach to verify proofs generated by \Metis prover to problems in
classical propositional logic.
To verify these \Metis proofs, we developed a tool able to translate
such derivations to \Agda proof-terms.
We formulate in type theory each inference rule found
in \Metis derivations. This allowed us to type-check \Metis proofs
step-by-step. We leave it for future work,
the opportunity to extend this work to other provers, proof assistants, and other formalisms.

\keywords{proof-reconstruction, type theory,
automatic theorem prover}

>>>>

\begin{abstract}
We describe a syntactical proof-reconstruction
approach to verify proofs generated by \Metis prover to problems in
classical propositional logic.
To verify these \Metis proofs, we developed a tool able to translate
such derivations to \Agda proof-terms.
We formulate in type theory each inference rule found
in \Metis derivations. This allowed us to type-check with \Agda, \Metis proofs
step-by-step. We leave it for future work,
the opportunity to extend this work to support other provers, proof assistants, and formalisms.

\keywords{proof-reconstruction, type theory, automatic theorem prover,
proof-assistant}

### 2017-10-20 paper/sections/introduction.tex
[ paper ] restructuring the paper :S.

Proof-reconstruction is a non-trivial task since it depends on the
integration of two complex systems.

On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories.
A \SAT solver (e.g., \name{zChaff}~\cite{Moskewicz2001} and
\name{MiniSat}~\cite{Een2004}) to prove unsatisfiability of \CNF
formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and
\name{DepQBF}~\cite{Lonsing2017}) to prove satisfiability and
invalidity of quantified Boolean formulas, a \SMT solver
(e.g \name{CVC4}~\cite{Barrett2011}, \name{veriT}~\cite{bouton2009},
and \name{Z3}~\cite{DeMoura2008}) to prove unsatisfiability of
formulas from first-order logic with theories, and a prover for
validity of formulas from first-order logic with equality
(e.g., \name{E}~\cite{Schulz:AICOM-2002},
\name{leanCoP}~\cite{Otten2008},
\Metis~\cite{hurd2003first},
\name{SPASS}~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II}~\cite{Benzmuller2008} and
\name{Satallax}~\cite{Brown2012}) or intuitionistic logic (e.g.,
\name{ileanCoP}~\cite{Otten2008},
\name{JProver}~\cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam},
\name{Coq}~\cite{coqteam},
\name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems i.e.,
operators, logic variables, axioms, and theorems.

A proof-reconstruction tool provides such an integration, translating
the reply generated by the prover into the formalism of the proof
assistant. Since the formalism of the source (the evidence
generated by the \ATP) is not necessarily the same logic in the
target, the reconstruction turns out to be a ``reverse engineering''
task. Then, reconstructing a proof involves an in-depth
understanding of the algorithms in the \ATP and the logic specification
in the \ITP. To begin with the proof-reconstruction, it is
necessary to have a proof-object from the \ATP in a consistent
format, that is, a full script describing step-by-step
with exhaustive details and without ambiguities, the derivation to
get the actual proof.
For problems in classical propositional logic (henceforth \CPL),
from a list of at least forty\footnote{\ATPs available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs (e.g.,
\name{CVC4}~\cite{Barrett2011}, \name{SPASS}, and
\name{Waldmeister}~\cite{hillenbrand1997}) and fewer replied with
a proof
in a file format like \TSTP~\cite{Sutcliffe-Schulz-Claessen-VanGelder-2006}
(e.g., \name{E}, \Metis, \name{Vampire}, and \name{Z3}),
\len{LFSC}~\cite{Stump2008}, or the \len{SMT-LIB}~\cite{Bohme2011} format.

@@

>>>>

Proof-reconstruction is a non-trivial task since it depends on the
integration between two complex systems.
We work in this study with one integration between an automatic theorem
prover (henceforth \ATP) and an interactive theorem prover (henceforth
\ITP).

The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems i.e.,
operators, logic variables, axioms, and theorems.

A proof-reconstruction tool provides such an integration, translating
the reply generated by the prover into the formalism of the proof
assistant.

% Since the formalism of the source (the evidence
% generated by the \ATP) is not necessarily the same logic in the
% target, the reconstruction turns out to be a ``reverse engineering''
% task.

Then, reconstructing a proof involves an in-depth
understanding of the algorithms in the \ATP and the logic specification
in the \ITP. To begin with the proof-reconstruction, it is
necessary to have a proof-object from the \ATP in a consistent
format, that is, a full script describing step-by-step
with exhaustive details and without ambiguities, the derivation to
get the actual proof.
For problems in classical propositional logic (henceforth \CPL),
from a list of at least forty\footnote{\ATPs available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs (e.g.,
\name{CVC4}~\cite{Barrett2011}, \name{SPASS}, and
\name{Waldmeister}~\cite{hillenbrand1997}) and fewer replied with
a proof
in a file format like \TSTP~\cite{Sutcliffe-Schulz-Claessen-VanGelder-2006}
(e.g., \name{E}, \Metis, \name{Vampire}, and \name{Z3}),
\len{LFSC}~\cite{Stump2008}, or the \len{SMT-LIB}~\cite{Bohme2011} format.

@@

The program and formalizations in \Agda described in this paper are available
as \verb!Git! repositories at \verb!Github!:

\begin{itemize}
  \item The \Athena program to translate \Metis proofs to \Agda code:\\
  \hfill\url{http://github.com/jonaprieto/athena}.\hfill
  \item A formal system in \Agda for propositional logic:\\
  \hfill\url{http://github.com/jonaprieto/agda-prop}.\hfill
  \item A formalization in \Agda to justify \Metis proofs in propositional
  logic:\\
  \hfill\url{http://github.com/jonaprieto/agda-prop}.\hfill
\end{itemize}

### 2017-10-20 paper/sections/metis-language-and-proofs.tex
[ paper ] restructuring the paper :S.

\Metis is an automatic theorem prover written in \len{ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. There are already some tools that perform proof-reconstruction
with \Metis. %Some of these systems are the following.

\name{Isabelle/HOL} incorporated \Metis as a macro-step
reconstruction tool for explaining proof steps (usually \CNF goals)
replied from other \ATPs like \name{CVC4}, \name{Vampire}, or
\name{Z3} (more details in Ref.~\cite{paulson2007source}).
\citeauthor{Farber2015} in \cite{Farber2015} present tactic based on \Metis for
\name{HOL Light} capable to challenge others tactics like \emph{MESON}
or the \emph{leanCoP} tactic~\cite{Farber2016}.

@@

\begin{figure}
\[
% \scalebox{0.9}{
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{axiom}
  \UnaryInfC{$C$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{assume $L$}
  \UnaryInfC{$L ∨ \neg L$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{refl $t$}
  \UnaryInfC{$t = t$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{$C$}
  \RightLabel{subst $\sigma$}
  \UnaryInfC{$\sigma\,C$}
\end{bprooftree}
\qquad
% }
\]
\[
% \scalebox{0.9}{
\begin{bprooftree}
  \AxiomC{}
  \RightLabel{equality $L$ $p$ $t$}
  \UnaryInfC{$\neg (L[p] = t) ∨ \neg L ∨ L[ p \mapsto t]$}
\end{bprooftree}
\qquad
\begin{bprooftree}
  \AxiomC{$L ∨ C$}
  \AxiomC{$\neg L ∨ D$}
  \RightLabel{resolve $L$}
  \BinaryInfC{$C ∨ D$}
\end{bprooftree}
% }
\]
\caption{CNF Inference rules of the \texttt{Metis} prover.}
\label{fig:metis-inferences}
\end{figure}

>>>>

\Metis is an automatic theorem prover written in \len{ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects
to justify its deduction steps from the basis of only six inference
rules.
In Fig.~\ref{fig:metis-inferences}, we show the three \Metis rules for
propositional logic.

\begin{figure}
\begin{equation*}
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{axiom}
    \UnaryInfC{$Γ ⊢ φ$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{assume $L$}
    \UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
  \end{bprooftree}
  \end{equation*}
  \vskip2mm
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ l ∨ φ$}
    \AxiomC{$Γ ⊢ ¬ l ∨ ψ$}
    \RightLabel{resolve $l$}
    \BinaryInfC{$Γ ⊢ φ ∨ ψ$}
  \end{bprooftree}
\end{equation*}
\caption{CPL inference rules of the \texttt{Metis} prover.}
\label{fig:metis-inferences}
\end{figure}

@@

### 2017-10-20 paper/sections/preliminaries.tex
[ paper ] restructuring the paper :S.

\subsection{Classical Propositional Logic}
\label{ssec:CPL}

@@

\emph{Derivations.}
A derivation of $φ$ is a mechanism to obtain the conclusion $φ$ from
the application of some deduction steps. Such deduction step has
the form of a set of propositional \emph{premises} that produces
propositions as the conclusion.

@@

\begin{figure}
\label{fig:derivation}
\begin{equation*}
\begin{bprooftree}
  \AxiomC{$φ$}
  \AxiomC{$ψ$}
  \RightLabel{∧-intro}
  \BinaryInfC{$φ ∧ ψ$}
\end{bprooftree}
\end{equation*}
\caption{Derivation example.}
\end{figure}

@@

\emph{Postulates.}
We only postulate the principle of excluded third
(\abbre{PEM}). Assuming \abbre{PEM}, we derive the
\emph{reductio ad absurdum} rule (\abbre{RAA}) as the
formulation of the principle of proof by contradiction, that is, a derivation
of a contradiction, $⊥$, from the hypothesis $(¬ φ)$, is a
derivation of $φ$.

\begin{equation*}
\label{eq:PEM-RAA}
\begin{bprooftree}
\AxiomC{}
\RightLabel{PEM,}
\UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
\end{bprooftree}\qquad
\begin{bprooftree}
\AxiomC{$Γ, ¬ φ ⊢ ⊥$}
\RightLabel{RAA.}
\UnaryInfC{$Γ ⊢ φ$}
\end{bprooftree}
\end{equation*}

\emph{Natural Deductions.}
The natural deduction designed by Gerhard Gentzen in 1930s defines a formal system for derivations that uses a set
of inference rules as the only way to obtain a conclusion.
We use this system to refer to formal proofs as natural deduction
proofs and conversely.
We define the \CPL formal system in
Fig.~\ref{fig:CPL-inference-rules}. The rules described in the aforementioned figure give us the complete
meaning of the connectives; how to introduce them and
how they can be remove them in the conclusion side of the theorems.

\begin{figure}
\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{}
\RightLabel{assume}
\UnaryInfC{$Γ , φ ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{weaken}
\UnaryInfC{$Γ , ψ ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{}
\RightLabel{$⊤$-intro}
\UnaryInfC{$Γ ⊢ ⊤$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ ⊥$}
\RightLabel{$⊥$-elim}
\UnaryInfC{$Γ ⊢ φ$}
\end{bprooftree}
%}
\]

\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ , φ ⊢ ⊥$}
\RightLabel{$¬$-intro}
\UnaryInfC{$Γ ⊢ ¬ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ ¬ φ$}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{$¬$-elim}
\BinaryInfC{$Γ ⊢ ⊥$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ$}
\AxiomC{$Γ ⊢ ψ$}
\RightLabel{$\wedge$-intro}
\BinaryInfC{$Γ ⊢ φ \wedge ψ$}
\end{bprooftree}
%}
\]

\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ \wedge ψ$}
\RightLabel{$\wedge$-proj$_1$}
\UnaryInfC{$Γ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ \wedge ψ$}
\RightLabel{$\wedge$-proj$_2$}
\UnaryInfC{$Γ⊢ ψ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{$\vee$-intro$_1$}
\UnaryInfC{$Γ ⊢ φ \vee ψ$}
\end{bprooftree}
%}
\]

\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ ψ$}
\RightLabel{$\vee$-intro$_2$}
\UnaryInfC{$Γ ⊢ φ \vee ψ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ , φ ⊢ \gamma $}
\AxiomC{$Γ , ψ  ⊢ \gamma$}
\RightLabel{$\vee$-elim}
\BinaryInfC{$Γ , φ \vee ψ ⊢ \gamma$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ , φ ⊢ ψ$}
\RightLabel{$\Rightarrow$-intro}
\UnaryInfC{$Γ ⊢ φ \Rightarrow ψ$}
\end{bprooftree}
%}
\]

>>>>

\subsection*{Classical Propositional Logic}
\label{ssec:CPL}

@@

\emph{Inference Rule.}
An inference rule is a deduction step of the form of a set of propositional
\emph{premises} that produces propositions as the conclusion.
A derivation is a

is a mechanism to obtain the conclusion $φ$ from
the application of some deduction steps. Such deduction step has

@@

\begin{figure}
\label{fig:derivation}
\begin{equation*}
\begin{bprooftree}
  \AxiomC{$φ ∧ ψ$}
  \RightLabel{∧-proj₂}
  \UnaryInfC{$ψ$}
  \AxiomC{$φ ∧ ψ$}
  \RightLabel{∧-proj₁}
  \UnaryInfC{$φ$}
  \RightLabel{∧-intro}
  \BinaryInfC{$ψ ∧ φ$}
\end{bprooftree}
\end{equation*}
\caption{Derivation example.}
\end{figure}

@@

% \emph{Postulates.}
% We postulate the principle of excluded third
% (\abbre{PEM}). Assuming \abbre{PEM}, we derive the
% \emph{reductio ad absurdum} rule (\abbre{RAA}) as the
% formulation of the principle of proof by contradiction, that is, a derivation
% of a contradiction, $⊥$, from the hypothesis $(¬ φ)$, is a
% derivation of $φ$.

% \begin{equation*}
% \label{eq:PEM-RAA}
% \begin{bprooftree}
% \AxiomC{}
% \RightLabel{PEM,}
% \UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
% \end{bprooftree}\qquad
% \begin{bprooftree}
% \AxiomC{$Γ, ¬ φ ⊢ ⊥$}
% \RightLabel{RAA.}
% \UnaryInfC{$Γ ⊢ φ$}
% \end{bprooftree}
% \end{equation*}

\emph{Natural Deductions.}
The natural deduction introduced by Gentzen in 1930s defines a formal system
for derivations that uses a set of inference rules as the only way to obtain a
conclusion. We use this system to refer to formal proofs as natural deduction
proofs and conversely.

In this document, we extend the \CPL formal system
Fig.~\ref{fig:CPL-inference-rules} described on
paper~\cite{Altenkirch2015} to include inferences rules for the biconditional
connective, $PEM$ and $RAA$ postulates, among others.

\begin{figure}
\begin{equation*}
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{assume}
    \UnaryInfC{$Γ , φ ⊢ φ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ φ$}
    \RightLabel{weaken}
    \UnaryInfC{$Γ , ψ ⊢ φ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{$⊤$-intro}
    \UnaryInfC{$Γ ⊢ ⊤$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ ⊥$}
    \RightLabel{$⊥$-elim}
    \UnaryInfC{$Γ ⊢ φ$}
  \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ , φ ⊢ ⊥$}
    \RightLabel{$¬$-intro}
    \UnaryInfC{$Γ ⊢ ¬ φ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ ¬ φ$}
    \AxiomC{$Γ ⊢ φ$}
    \RightLabel{$¬$-elim}
    \BinaryInfC{$Γ ⊢ ⊥$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ φ$}
    \AxiomC{$Γ ⊢ ψ$}
    \RightLabel{$\wedge$-intro}
    \BinaryInfC{$Γ ⊢ φ \wedge ψ$}
  \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ φ \wedge ψ$}
    \RightLabel{$\wedge$-proj$_1$}
    \UnaryInfC{$Γ⊢ φ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ φ \wedge ψ$}
    \RightLabel{$\wedge$-proj$_2$}
    \UnaryInfC{$Γ⊢ ψ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ φ$}
    \RightLabel{$\vee$-intro$_1$}
    \UnaryInfC{$Γ ⊢ φ \vee ψ$}
  \end{bprooftree}
\end{equation*}

\begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ ψ$}
    \RightLabel{$\vee$-intro$_2$}
    \UnaryInfC{$Γ ⊢ φ \vee ψ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ , φ ⊢ γ $}
    \AxiomC{$Γ , ψ  ⊢ γ$}
    \RightLabel{$\vee$-elim}
    \BinaryInfC{$Γ , φ \vee ψ ⊢ γ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ , φ ⊢ ψ$}
    \RightLabel{$\Rightarrow$-intro}
    \UnaryInfC{$Γ ⊢ φ \Rightarrow ψ$}
  \end{bprooftree}
\end{equation*}
